A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-03 paint-contour-tree.tex
8.4 added

\begin{theorem} \label{thm:main-alg} Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}

@@

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
one may think that that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
of building the join trees can be related to the cost of building $\cC(\MM)$.

@@

The key idea is \emph{paint spilling}. Start with each maxima having a large can of paint, with distinct
colors for each maxima. In arbitrary order, we spill paint from each maxima, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
paint the interior of facets.} Alternately, this process is restricted to the 1-skeleton of $\NN$.
Furthermore, our paints do not mix, so each edge receives a unique color, which is decided by the first
paint to reach it.

@@

Note that a painting only colors edges, not vertices. Our definition also does not require the timing aspect
of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maxima
colors are distinct. A few comments on paint vs water. The interface between two regions of different color
is \emph{not} a contour, and is actually difficult to characterize. So we cannot apply the divide-and-conquer
approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
dominant input, using water to divide the input $\NN$ raise the complexity too much.

Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to \Fig{colors}.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
(So ignore the yellow mound in \Fig{colors}.)
Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
So this indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.

This is precisely how we hope to exploit the information in the painting. We prove later that
when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
The merging of the colors is not explicitly as that would be too expensive; we maintain a union-find
data structure for that.

Of course, things are more complicated when there are other mounds. There may be a yellow mound,
corresponding to $\ell_3$ that joins with the blue mound higher up (at vertex $u$). In $\jc(\MM)$, $\ell_2$ and $\ell_3$
are sibling leaves, and $\ell_1$ is a sibling of some ancestor of these leaves. So we cannot
merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
We know that $u$ must also be touched by blue. So all critical vertices touched by blue
are put into a priority queue keyed by height, and vertices are handled in that order.

\begin{figure}[h]\centering
    \includegraphics[width=.28\linewidth]{figs/mound}%
    \hspace{.1in}
    \includegraphics[width=.28\linewidth]{figs/mound2}%
    \hspace{.1in}
    \includegraphics[width=.28\linewidth]{figs/mound3}%
    \hspace{.1in}
    \includegraphics[width=.065\linewidth]{figs/moundTree}
    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.
    On the far right, the contour tree.}
    \label{fig:colors}
\end{figure}

@@

The cost of heap operations depends on the sizes of the heaps, which keeps changing because
of the repeated merging. This is a major headache for the analysis. There are situations where
the initial heap sizes are small, but they eventually merge to create larger heaps. 
We employ a variant
of \emph{heavy path decompositions}, first used for Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
The final analysis is basically charges expensive heap operations to long paths in the decomposition.

@@

We have a fairly simple construction that works for terrains. In $P(T)$, consider the path
$p$ that involves the root. The base of the construction is a conical ``tent", and there
will be $|p|$ triangular faces that will each have a saddle. The heights of these saddles
can be varied arbitrarily, and that will give $|p|!$ different choices. Each of these saddles
will be connected to a recursive construction involving other paths in $P(T)$. Effectively,
one can think of tiny tents that are sticking out of each face of the main tent. The contour
trees of these tiny tents attach to a main branch of length $|p|$. Working out the details,
we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. It is possible to argue
that our algorithm requires $\Theta(\cost(P(T)))$ on all these instances.

@@

\Ben{This is already defined in section 3, should we use the same numbering?}
\begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$.

@@

\begin{theorem}
 Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\jc(\MM)$.
\end{theorem}
\begin{proof}
 First observe that each vertex can be processed at most once by $\build(\MM)$.  By \Cor{terminate}, we know that as long as there 
 is an unprocessed vertex, $\update(K)$ will be called and will terminate with a ripe vertex which is ready to be processed. 
 Therefore, eventually all vertices will be processed, and so by \Clm{process} the algorithm will terminate having 
 computed $\jc(\MM)$.
\end{proof}

@@

Specifically, proving the above theorem will imply our main result, \Thm{main-alg}, which we now restate and prove.

\begin{theorem}
Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}
\begin{proof}
\Ben{Prove the theorem, and change the numbering on the theorem.}
\end{proof}

@@

>>>>

\begin{theorem} \label{thm:main-alg} Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is an algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}

@@

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
one may think that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
of building the join tree can be related to the cost of building $\cC(\MM)$.

@@

The key idea is \emph{paint spilling}. Start with each maxima having a large can of paint, with distinct
colors for each maxima. In arbitrary order, we spill paint from each maxima, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
paint the interior of facets.} That is, this process is restricted to the 1-skeleton of $\NN$.
Furthermore, our paints do not mix, so each edge receives a unique color, which is decided by the first
paint to reach it.

@@

Note that a painting only colors edges, not vertices. Our definition also does not require the timing aspect
of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maxima
colors are distinct. A few comments on paint vs water. The interface between two regions of different color
is \emph{not} a contour, and is actually difficult to characterize. So we cannot apply the divide-and-conquer
approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
dominant input, using water to divide the input $\NN$ raises the complexity too much.

Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
So this indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.

This is precisely how we hope to exploit the information in the painting. We prove later that
when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
The merging of the colors is not explicitly performed as that would be too expensive; we maintain a union-find
data structure for that.

Of course, things are more complicated when there are other mounds. There may be a yellow mound,
corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\jc(\MM)$, $\ell_1$ and $\ell_3$
are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
We know that $u$ must also be touched by blue. So all critical vertices touched by blue
are put into a priority queue keyed by height, and vertices are handled in that order.

\begin{figure}[h]\centering
    \includegraphics[width=.25\linewidth]{figs/redBlue}%
    \hspace{.07in}
    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
    \hspace{.07in}
    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
    \hspace{.09in}
    \includegraphics[width=.3\linewidth]{figs/mound}%
    \hspace{.07in}
    \includegraphics[width=.065\linewidth]{figs/moundTree}
    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
    \label{fig:colors}
\end{figure}

@@

The cost of heap operations depends on the sizes of the heaps, which keeps changing because
of the repeated merging. This is a major headache for the analysis. There are situations where
the initial heap sizes are small, but they eventually merge to create larger heaps. 
We employ a variant
of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
The final analysis basically charges expensive heap operations to long paths in the decomposition.

@@

We have a fairly simple construction that works for terrains. In $P(T)$, consider the path
$p$ that involves the root. The base of the construction is a conical ``tent", and there
will be $|p|$ triangular faces that will each have a saddle. The heights of these saddles
can be varied arbitrarily, and that will give $|p|!$ different choices. Each of these saddles
will be connected to a recursive construction involving other paths in $P(T)$. Effectively,
one can think of tiny tents that are sticking out of each face of the main tent. The contour
trees of these tiny tents attach to a main branch of length $|p|$. Working out the details,
we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. It is possible to argue
that our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.

@@

We now restate \Def{paint}.

\begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$.

@@

\begin{theorem}
\label{thm:correct}
 Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\jc(\MM)$.
\end{theorem}
\begin{proof}
 First observe that each vertex can be processed at most once by $\build(\MM)$.  By \Cor{terminate}, we know that as long as there 
 is an unprocessed vertex, $\update(K)$ will be called and will terminate with a ripe vertex which is ready to be processed. 
 Therefore, eventually all vertices will be processed, and so by \Clm{process} the algorithm will terminate having 
 computed $\jc(\MM)$.
\end{proof}

@@

We conclude this section in \Sec{implications} by showing that proving this theorem implies our main result \Thm{main-alg}.

@@

\subsection{Our Main Result}
\label{sec:implications}

We now show that \Thm{runtime} allows us to upper bound the running time for our join tree and contour 
tree algorithms in terms of path decompositions.

\begin{theorem}
Consider a PL-Morse $f:\MM \mapsto \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$. 
There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$.
\end{theorem}
\begin{proof}
By \Thm{correct} we know that $\build(\MM)$ correctly outputs $\jc(\MM)$, and by \Lem{runTimeUpper}
we know this takes $O(\sum_{v \in \jc(\MM)} \log |H_v| + t\alpha(t) + N)$ time, where the $H_v$ values are determined as in \Def{initialColoring}.
Therefore by \Thm{runtime}, $\build(\MM)$ takes $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$ 
time to correctly compute $\jc(\MM)$.
\end{proof}

This result for join trees easily implies our main result, \Thm{main-alg}, which we now restate and prove.

\begin{theorem}
Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $\cC=\cC(\MM)$
has maximum degree $3$. There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}
\begin{proof}
First, lets review the various pieces of our algorithm. 
On a given input simplicial complex, we first we break it into extremum dominant pieces using $\rain(\MM)$ (and in $O(|\MM|)$ time by \Thm{rain-time}).  
Specifically, \Lem{rain-1} proves that the output of $\rain(\MM)$ is a set of extremum dominant pieces, 
$\MM_1, \dots, \MM_k$, and \Clm{rain-reeb} shows that given the contour trees, $\cC(\MM_1), \dots, \cC(\MM_k)$ , 
the full contour tree, $\cC(\MM)$, can be constructed (in $O(|\MM|)$ time).

Now one of the key observations was that for extremum dominant manifolds, computing the contour tree is roughly the same as 
computing the join tree.  Specifically, \Thm{contour-tree} implies that given $\jc(\MM_i)$ , 
we can obtain $\cC(\MM_i)$ by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
However, by the above theorem we know $\jc(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
(where $t_i$ and $N_i$ are the number of critical points and facets when restricted to $\MM_i$).

At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\jc(\MM_i))$.
Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
compute the contour tree in time
\[
O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i} = O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
\]

%\Thm{rain-time} proves that $\rain$ takes $O(|\MM|)$ time

%\Lem{rain-1} proves that output of $\rain$ are extemum dominant.

%\Clm{rain-reeb} shows that given the contour trees of the output manifolds of $\rain$, the full contour tree can be constructed in $O(|\MM|)$ time:

%\Thm{contour-tree} implies that given an $\jc(\MM)$ for an etremum dominant $\MM$, we can obtain $\cC(\MM)$ in by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time). 
\end{proof}
